---
title: "clockchart-construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clockchart-construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clockplot)
```

This document reveals what's going on behind the scene--how the charts are 
constructed. For the details on how to use the package, see [clockplot](clockplot.html).

Broadly speaking, the construction of the clockcharts have two steps. First, 
the skeleton, the clock on which event times are plotted is constructed. For 
this, we need a circle, on which we have to mark 24 dots to denote 24 hours, 
starting from 00:00 (12 AM) to 23:00 (11 PM). 

It is enough to draw a unit circle, although a circle of any radius would do. 
In `R`, it is easiest to draw a circle using the complex equation. For a unit
circle, the equation is $e^{i\theta}$, where $e$ is Euler's number, an 
irrational constant having the approximate value of 2.71828; $i$ stands for 
imaginary unit, equal to $\sqrt{-1}$; and $\theta$ refers to the trigonometric
angle (in radian?). 


