---
title: "clockchart-construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clockchart-construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*This is only for those who want to know how the idea came into being and got implemented.* You are NOT required to read it as part of using the package. 

This document reveals what's going on behind the scene--how the charts are 
constructed--how the idea was conceived and implemented after numerous trial
and error. For the details on how to use the package, see [clockplot](clockplot.html).

<span style = "color:Orange;">⚠</span> The plots shown here may not produce the
expected output due to the behavior of Rmarkdown. To see the actual appearance,
run the codes given. 


Broadly speaking, the construction of the clockcharts have two steps. 

i. Construction of the 24-hour clock
ii. Plotting event times on the clock

Each steps has two substeps:

a. The mathematical idea
b. The programming

### The Clock Skeleton

First we need to construct the skeleton, the clock on which event times are 
plotted. For this, we need a circle, on which we have to mark 24 
dots to denote 24 hours, starting from 00:00 (12 AM) to 23:00 (11 PM). 

It is enough to draw a unit circle, although a circle of any radius would do. 
In `R`, it is easiest to draw a circle using the complex equation. For a unit
circle, the equation is $e^{i\theta}$, where $e$ is Euler's number, an 
irrational constant having the approximate value of 2.71828; $i$ stands for 
imaginary unit, equal to $\sqrt{-1}$; and $\theta$ refers to the trigonometric
angle in radian. To make a complete circle, we need to allow $\theta$ to
take values from 0 to 360 degrees ($2\pi$ in radian). 

Programmatically, we need to create some angles ranging from 0 to $2\pi$ and 
then connect them to make a complete circle. Let us create 100 angles and use
them in the circle equation ($e^{i\theta}$). The trigonometric form of the 
equation is $sin^2 \theta + cos^2 \theta = 1$, while the algebraic form is
$x^2+y^2 = 1$. The complex form, however, is more convenient. 

```{r, fig.height=4, fig.width=4, fig.align='center', fig.cap="The Unit Circle"}
k <- 100
timepoint <- exp(1i * 2 * pi * (k:1) / k)
plot(timepoint, pch = 19, type = "b")
```

Here `R` just takes the complex numbers and plots the values of the real part
on the X axis and the corresponding values of the imaginary part on the Y axis. 

Now we have to mark the hours on the circle to make it look like a clock. We 
have to show 24 hours marks as well some marks for minute marks, corresponding 
to the fraction of the hour. This clock is slightly different from a traditional
clock not only in that it has 24 hours instead of 12, but also that it has 
only one hand, the hour hand, with minute part added to the hour mark as the 
fraction; the second part, having minor contribution, is ignored. 

First, let us see the output with 24 hour marks. 

```{r circle, fig.height=4, fig.width=4, fig.align='center'}
k <- 24
hour <- exp(1i * 2 * pi * (k:1) / k)
plot(hour, pch = 19)
```

At this time, we move to the `R` package `ggplot2` to produce better graphics. 
We need to create an `R` data frame for plotting ggplots. We also mark some 
points for minutes. 

First, let us create the required data frame

```{r, fig.height=4, fig.width=4, fig.align='center'}
k <- 24 # Hours
subk <- 24 * 4 # Fraction of hours
times <- exp(1i * 2 * pi * (k:1) / k)
subtimes <- data.frame(SubT = exp(1i * 2 * pi * (subk:1) / subk))
ampm <- c(rep(" AM", 6), rep(" PM", 12), rep(" AM", 6))
library(tibble)
dfclock <- tibble(
  time = times,
  hour = c(6:12, 1:12, 1:5), # May not be needed
  label = paste0(c(6:12, 1:5), ampm)
)
```

**Explanation**

`k = 24` indicates there are 24 hours. Then in each hour is divided into 4 parts
, corresponding to the elapse of successive 15 minutes. Thus, a small point on 
\@ref(fig:clock24) in the middle of two larger point (hour) means half an hour
has elapsed. For instance, the the point in the middle of 7 PM and 8 PM is 7:30 
PM. 

Now the code `exp(1i * 2 * pi * (k:1) / k)` gives us all the angles 
corresponding to 24 hours on the clock. Here, we have to note that the angles
in trigonometry starts from $0^o$ (X-axis) and rotates counter-clockwise, the 
direction opposite to the rotation of clock hands (the significance of the
difference will be seen shortly). But the clock moves clockwise. 
At 0 degree, we have 3 in a 12-hour clock and 6 in 24-hours clock.

Similarly, the code `SubT = exp(1i * 2 * pi * (subk:1) / subk)` gives us the
angles corresponding to the fraction of an hour. As we have already mentioned, 
the clock hand rotates clock-wise, starting from the $0^o$ angle, so we name 
the labels 6 PM to 11 PM, then we have 12 PM to 11 PM, followed by 12 AM to 
5 AM. 

The code `c(rep(" AM",6), rep(" PM",12), rep(" AM",6))` ensures this sequence. 

We then place the data into two data frames `subtimes` and `dfclock`, since
they are unequal dimensions. 

```{r skeleton, fig.height=4, fig.width=4, fig.align='center'}
library(ggplot2) # Load the package ggplot2
p1 <- dfclock %>% ggplot() +
  geom_path(data = subtimes, aes(Re(SubT), Im(SubT))) +
  # Connect Last two missing points
  geom_line(data = subtimes %>% dplyr::slice(-c(2:95)), aes(Re(SubT), Im(SubT))) +
  theme(aspect.ratio = 1) +
  geom_text(data = dfclock, aes(Re(time) * 1.1, Im(time) * 1.1, label = label)) +
  geom_point(
    data = subtimes,
    aes(Re(SubT), Im(SubT)), shape = 19, color = "black", size = 0.6
  ) +
  geom_point(aes(Re(time), Im(time)), color = "black", size = 1.8)
p1
```

Next, we plot the lines using the `ggplot2` package. The functions `geom_path()`
does the job, while `geom_line()` is used to fill the gap between the first and 
the last point. `aspect.ratio = 1` is set to ensure the chart size do not 
depend on resizing of `Rstudio` Viewer size.

The points corresponding to the hours are labelled AM or PM using the 
`geom_text()` function. We do not need the axis labels and ticks. Let us 
remove them:

```{r skeleton_clean, fig.height=4, fig.width=4, fig.align='center'}
p1 + theme(
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank()
) +
  labs(x = "", y = "")
```

This completes our first part of the plot; we get the skeleton of the clock. 

The second and final part of the task involves plotting event times on the
appropriate points. Now, this task has several variations. In the simplest 
form, we just have a chart, so called `clock_chart`, which plots the event
times on the 24-hour clock. An optional argument `Col` can be used to change
the color of the clock hands. The other variations will be discussed later, when
we complete the discussion on how the event times are prepared for plotting.

#### Preparation of Event Times Data

We have data in the format `HH:MM:SS`. We need to convert this data to `(x,y)`
coordinates. In the unit circle, for example, the 12 AM should be plotted on
the Y-axis (x = 0, y = 1) and 6 AM on the X axis (x = 1, y = 0), although, to
avoid plotting on the circumference, we decrease the length a little (more on
this later). 

To convert the `HH:MM:SS` to coordinates of the unit circle, we first separate 
the parts of time into `HH`, `MM`, and `SS`. We accomplish that using the 
`dplyr` and `tidyr` package. 

```{r}
# Data to check on
chkdf <- data.frame(
  time = c("06:00:00", "12:00:00", "17:30:00", "00:05:25"),
  value = sample(10, 4)
)
library(dplyr)
chkdf %>% tidyr::separate_wider_delim(
  cols = time,
  names = c("hour", "minute", "second"),
  cols_remove = FALSE,
  delim = ":"
)
```

This gives us `hour`, `minute`, and `second` as character (`chr`) vectors, but 
we need them as numeric vectors. The conversion is made with `as.numeric` 
function. 

Here the challenge now is to combine the hour and the minute into one singe
time, which we would later convert to coordinate. We intentionally ignore the
second (`SS`), since it has too negligible contribution on the plot.

Here a short algorithm is to be used:

Note here that 30 minutes is equivalent to half an hour, so `1:30:00` corresponds
to 1.5 hour. We need to recall that time values range from 0 to 60, while 
our decimal number system gives us the suitable range 1 to 100. In our number
system, reaching 100 indicates we have completed some work, while clockworks 
say reaching 60 completes one cycle of the task. 

The hour part does not need conversion. 23 is 23 and 22 is 22. 

Thus, 15 minutes on the clock says $\frac{15}{60} = \frac14$ th or $25\%$ work, 
30 means $\frac{30}{60} = \frac12 = 0.5$ or $50\%$ work, and 50 means 
$\frac{50}{60}$ or $83\%$ work have been completed. 

The task is further complicated due to the clock and angles rotating in
opposite directions.

In trigonometry, angles start at 0° on the positive X-axis and increase
counter-clockwise. But a clock starts at 12 o'clock (positive Y-axis) and
moves clockwise. To reconcile this, we need a coordinate transformation.

### The Complete Coordinate Conversion Algorithm

We convert `HH:MM:SS` to `(x, y)` coordinates on the unit circle as follows:

1. **Parse time to fractional hours since midnight**
   - `timc = hour + minute/60` (seconds are ignored as visually negligible)
   - Example: `17:30` → `17.5` hours

2. **Map fractional hours to radians (clockwise from 12 o'clock)**
   - The clock has 24 hours = 2π radians
   - 12 AM (midnight) = 0 hours = π/2 radians (pointing up)
   - 6 AM = 6 hours = 0 radians (pointing right)
   - 12 PM = 12 hours = 3π/2 radians (pointing down)
   - 6 PM = 18 hours = π radians (pointing left)
   - Formula: `time_angle = (6 - timc) * π/12` for `0 ≤ timc ≤ 6`
   - Formula: `time_angle = (30 - timc) * π/12` for `6 < timc ≤ 24`
   - This piecewise formula handles the clock's discontinuity at 6 AM/6 PM

3. **Convert to Cartesian coordinates**
   - `x = r * cos(time_angle)`
   - `y = r * sin(time_angle)`
   - Where r = 0.95 for event points, r = 1.0 for clock skeleton
   - The segment from origin `(0, 0)` to `(x, y)` forms the "clock hand"

This algorithm is implemented in `conv_data()`, `conv_data_col()`, and `conv_data_len()` in `R/conv_data.R`.

### Mapping to `ggplot2` Aesthetics

Once we have `(x0, y0, x1, y1)` for each event (where `(x0, y0) = (0, 0)` and `(x1, y1)` is the tip of the clock hand), the chart functions map these to `ggplot2` geoms:

| Function | `geom_segment` | `geom_point` | Color | Size |
|----------|----------------|--------------|-------|------|
| `clock_chart()` | Fixed color | Fixed color | `Col` arg | Constant |
| `clock_chart_col()` | Mapped to `crit` | Mapped to `crit` | Gradient `high`/`low` | Mapped to `crit` |
| `clock_chart_qnt()` | Mapped to `len` (color) | Mapped to `len` (color), `Col` (size) | Gradient `high`/`low` | Mapped to `Col` or `len` |
| `clock_chart_qlt()` | Mapped to `crit` (factor) | Mapped to `crit` (factor) | Discrete (brewer) | Constant |

The `basic_clock()` function (in `R/basic_clock.R`) builds the static clock face:
- Circle via `geom_path` on 96 points (24 hours × 4 quarters)
- Hour labels via `geom_text` at radius 1.1
- Hour ticks via `geom_point` at radius 1.0
- Quarter-hour ticks via `geom_point` at radius 1.0, size 0.6
- `theme(aspect.ratio = 1)` + blank axes for a clean clock face

### Cyclic and Period Charts: Different Geometry

The cyclic charts (`cyclic_chart()`, `day_chart()`, `week_chart()`, `year_chart()`) use a fundamentally different approach: **polar bar charts** (`coord_polar()`) rather than clock hands.

- `cyclic_chart()`: General-purpose; maps any periodic variable (hours, days, months) to angle via `coord_polar()`, values to radius via `geom_col()`
- `day_chart()`: 24-hour rose plot; fixed hour labels (6 AM–5 AM); `geom_col` with gradient fill
- `week_chart()`: 7-day rose plot; fixed day labels (Saturday–Friday); `geom_col` with gradient fill
- `year_chart()`: 12-month rose plot; `month.name` labels; `geom_col` with gradient fill
- `plan_day()`/`plan_week()`: Categorical planning tools; `geom_col` with `fill = activity` + `geom_text` for labels; uses `scale_fill_brewer("Set2")`

These do **not** use the complex-number coordinate conversion. They rely entirely on `ggplot2::coord_polar()` to map linear positions to angles.

### Design Decisions and Rationale

| Decision | Rationale |
|----------|-----------|
| **24-hour clock** (not 12-hour) | Avoids AM/PM ambiguity; natural for timestamps; matches `HH:MM:SS` format |
| **Ignore seconds** | Visual angle change < 0.1°; adds noise without insight |
| **Complex numbers for skeleton** | `exp(1i * θ)` elegantly generates circle points; simpler than `sin`/`cos` loops |
| **Hand length ∈ [0.5, 0.95]** | `0.5` avoids origin clutter; `0.95` leaves room for labels at 1.1 |
| **Single hand (hour + minute fraction)** | Traditional clocks have two hands; here minute is encoded as fractional hour position — cleaner for dense events |
| **`coord_polar()` for cyclic charts** | Standard ggplot2 pattern; leverages existing polar coordinate system |
| **Factor warning at >5 categories** | Human color discrimination limit; suggests `clock_chart()` instead |




