---
title: "clockchart-construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clockchart-construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document reveals what's going on behind the scene--how the charts are 
constructed. For the details on how to use the package, see [clockplot](clockplot.html).

<span style = "color:Orange;">⚠</span> The plots shown here do not produce
expected output due to the behavior of Rmarkdown. To see the actual appearance,
run the codes given. 


Broadly speaking, the construction of the clockcharts have two steps. 

i. Construction of the 24-hour clock
ii. Plotting event times on the clock

Each steps has two substeps:

a. The mathematical idea
b. The programming

### The Clock Skeleton

First we need to construct the skeleton, the clock on which event times are 
plotted. For this, we need a circle, on which we have to mark 24 
dots to denote 24 hours, starting from 00:00 (12 AM) to 23:00 (11 PM). 

It is enough to draw a unit circle, although a circle of any radius would do. 
In `R`, it is easiest to draw a circle using the complex equation. For a unit
circle, the equation is $e^{i\theta}$, where $e$ is Euler's number, an 
irrational constant having the approximate value of 2.71828; $i$ stands for 
imaginary unit, equal to $\sqrt{-1}$; and $\theta$ refers to the trigonometric
angle in radian. To make a complete circle, we need to allow $\theta$ to
take values from 0 to 360 degrees ($2\pi$ in radian). 

Programmatically, we need to create some angles ranging from 0 to $2\pi$ and 
then connect them to make a complete circle. Let us create 100 angles and use
them in the circle equation ($e^{i\theta}$). The trigonometric form of the 
equation is $sin^2 \theta + cos^2 \theta = 1$, while the algebraic form is
$x^2+y^2 = 1$. The complex form, however, is more convenient. 

```{r}
k <- 100
timepoint <- exp(1i * 2 * pi * (k:1) / k)
plot(timepoint, pch = 19)
```

Here `R` just takes the complex numbers and plots the values of the real part
on the X axis and the corresponding values of the imaginary part on the Y axis. 

Now we have to mark the hours on the circle to make it look like a clock. We 
have to show 24 hours marks as well some marks for minute marks, corresponding 
to the fraction of the hour. This clock is slightly different from a traditional
clock not only in that it has 24 hours instead of 12, but also that it has 
only one hand, the hour hand, with minute part added to the hour mark as the 
fraction; the second part, having minor contribution, is ignored. 

First, let us see the output with 24 hour marks. 

```{r}
k <- 24
hour <- exp(1i * 2 * pi * (k:1) / k)
plot(hour, pch = 19)
```

At this time, we move to the `R` package `ggplot2` to produce better graphics. 
We need to create an `R` data frame for plotting ggplots. We also mark some 
points for minutes. 
