---
title: "clockplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clockplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clockplot)
```

The clockpolot offers three types of plot

1. Clock chart
2. Period chart
3. Period planner

The first shows event times (HH:MM:SS) on a 24 hour clock. The times can be arbitrary and not necessarily specific hours at all. 

The second type allows you to show values during a certain duration (day, week, month, and year)

Finally, the third one can help you make a plan in a certain period (day, week, month, and year), presenting tasks against period. 

Will create another vignette to explain how it all was made.

## Examples 

Let us see examples of each of the function. 

### clock_chart()

This creates a clockplot, showing event times on a 24-hour clock. The hands are
neither colored nor length modified by a criteria. To do so, please use other 
`clock_chart*()` functions (see below).

To make use of this simple chart, we need a data frame with at least one column
showing the times of a certain event. The format should be in HH:MM:SS (for
example 19:50:10). If you have times in HH:MM format, use `make_hms` function to
convert to HH:MM:SS format. 

First, let's create a sample data.

```{r clock-chart, fig.align = 'center'}
library(tibble) # to make a tibble (tidy data frame)
df <- tibble::tibble(hr = sample(0:23, 30, replace = TRUE), # 30 values
                  mnt = sample(0:59, 30, replace = TRUE),
                  sec = sample(0:59, 30, replace = TRUE),
                  time = paste(hr, mnt, sec, sep = ":"),
                  value = sample(40,30))
clock_chart_len(df, time, len = value, col = "blue")
```


